---
title: "The Effects of Welfare Reform on Child Poverty in the UK"
author: "Matteo Larrode"
date: "`r Sys.Date()`"
output: md_document
---
# The Effects of Welfare Reform on Child Poverty in the UK

In this dissertation, I conduct a quantitative analysis of the effect of welfare policy changes in the UK on the proportion of children living in relative low-income families. 

This Markdown includes the data analysis component of this dissertation. It will first include some exploratory data analysis, and then will be used to run the other components of the quantitative analysis of the impact of UK benefit policies on the number of children living in relative low income families. Note that the collection and cleaning of the data that will be used here is done in scripts in the `data-raw` folder of this repository, and the datasets are available in the `data` folder. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r package-loading, include=FALSE}
library(tidyverse)
library(fixest)

# pkgload::load_all(".")
```

## Data

### Outcome: Children Living in Relative Low Income Families

Data is downloaded from [Stat-Xplore](https://stat-xplore.dwp.gov.uk/webapi/jsf/login.xhtml), the Department of Work and Pensions (DWP) data distribution platform.

- Low income is a family whose equivalised income is below 60 per cent of median household incomes. For Absolute low income involves, the median of the 2010/11 year is used. For Relative low income, the comparison is made to the median of the current year. This dissertation focuses on Relative low income because it better adjusts for economic changes. Relative income is Before Housing Costs (BHC) and includes contributions from earning, state support and pensions. Equivalisation adjusts incomes for household size and composition.

- A family must have claimed Child Benefit and at least one other household benefit (Universal Credit, tax credits or Housing Benefit) at any point in the year to be classed as low income in these statistics

- Children are defined as dependent individuals aged under 16; or aged 16 to 19 in full-time non-advanced education.

Visit the dedicated [DWP page](https://www.gov.uk/government/publications/children-in-low-income-families-local-area-statistics-background-information-and-methodology/background-information-and-methodology-children-in-low-income-families-local-area-statistics) for further information. 

#### Relative measure

To get the **proportion** of children in low income families, I used [yearly population estimates](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/estimatesofthepopulationforenglandandwales) provided by the Office for National Statistics (for England and Wales), and [time series data](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/population/population-estimates/mid-year-population-estimates/population-estimates-time-series-data) from the National Records of Scotland. Because not all individuals aged 16 to 19 years old are counted as children in the CiRLIF measure, population counts are summed for individuals aged 0-15 years old, and used to divide the absolute measure of CiRLIF provided by the DWP.

**Geography: Local authorities (called ltla for Lower Tier Local Authorities)**

- England: English local authority districts (309)
- Wales: Unitary authorities (22)
- Scotland: Scottish council areas (32)
- Northern Ireland: District council areas (11)

### Main Independent Variable: Households on Universal Credit

Data for the number of households who have a calculated entitlement for Universal Credit is downloaded from [Stat-Xplore](https://stat-xplore.dwp.gov.uk/webapi/jsf/login.xhtml), the Department of Work and Pensions (DWP) data distribution platform.

Data is also available regarding the family type, number and age of children, and work status of the household claiming Universal Credit. For further information, visit the dedicated [DWP page](https://www.gov.uk/government/publications/universal-credit-statistics-background-information-and-methodology/universal-credit-statistics-background-information-and-methodology).

#### Relative measure

To calculate the proportion of households on Universal Credit for each local authority, numbers of households on UC were divided by historical annual estimates of the number of households published by the Office for National Statistics: [2004-2019](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/families/adhocs/13586estimatednumberofhouseholdsbyselectedhouseholdtypeslocalauthoritiesinenglandandwalescountiesandregionsofenglandscottishcouncilareasandgreatbritainconstituentcountries2004to2019), [2020](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/families/adhocs/14432estimatednumberofhouseholdsingreatbritaininnuts1nuts3englishandwelshlocalauthoritiesandscottishcouncilareas2020), [2021](https://www.ons.gov.uk/datasets/TS041/editions/2021/versions/3).

**Geography: Local authorities (called ltla for Lower Tier Local Authorities)**

- England: English local authority districts (309)
- Wales: Unitary authorities (22)
- Scotland: Scottish council areas (32)
- *[Data for Northern Ireland](https://www.communities-ni.gov.uk/articles/universal-credit-statistics) is made available by the Northern Ireland Department for Communities, but not yet collected in this package*

#### Universal Credit Rollout: Binary variable

The official [UC 'Full Service' transition rollout schedule](https://assets.publishing.service.gov.uk/media/5ab507c8e5274a1aa2d414d1/universal-credit-transition-rollout-schedule.pdf) published by the DWP was used to create explanatory variables measuring cross-local authority variation in the timing of the UC rollout (Hardie, 2023: 1035). The first is a binary variable indicating whether UC ‘Full Service’ had rolled out yet in each year in each local authority in the sample (coded: 0 = ‘no', 1 = ‘yes’). To mitigate bias and ensure robustness in the conversion of the monthly Universal Credit rollout data to annual data for the analysis, I coded multiple variables with different thresholds — December of the previous year, April, and October — to determine the minimum duration of Universal Credit presence within a financial year required for a local authority to be classified as having implemented UC. This first binary variable allows to study the staggered rollout of UC as a natural experiment, with a delayed treatment across local authorities.

The 'Full Service' iteration of Universal Credit was chosen for analysis because it represents the comprehensive form of UC, making it accessible to all types of claimants, including those with housing expenses. The initial 'Live Service' variant was not considered in this study due to its limited scope, primarily involving claims from single, unemployed individuals without housing cost claims, thereby having minimal impact on children from low-income families. (Hardie, 2023: 1035). 

#### Universal Credit Rollout Duration: Categorical variable

If a significant effect of the rollout on children living in low income families in indeed observed, this effect could be expected to vary depending on how long UC has been rollout out in the local authority. One possible scenario is that the effect fades over time because the administrative system adapts. Conversely, as time goes on, more people are reached by Universal Credit, and the effect on children in low income families could become larger. For the purpose of studying the long-term effect of the rollout, we build a secondary independent variable ‘UC Full Service Rollout by Rollout Length’; a categorical variable indicating whether UC ‘Full Service’ had rolled out yet in each month in each local authority , and if so for how long (coded: 0 = ‘No (pre-rollout)’ 1 = ‘Yes (1-6 months post)’, 2 = ‘Yes (7-12 months post)’, 3 = ‘Yes (13-24 months post)’, and 4 = 'Yes(25+ months post)') (Hardie, 2023: 1035). The proportion of children in low income families is measured for a financial year, so the number of months is measured from the month of the rollout to the middle of the financial year (October). 

##  I) Relationship Between Households on UC and Children in Low Income Families

I start with an examination of the relationship between the rollout of Universal Credit, more precisely the proportion of households on Universal Credit, and the proportion of children living in low income household. 

I join the datasets of households on UC and children living in low income, on local authority and year.

```{r message=FALSE}
data("children_low_income_ltla")
data("UC_households_ltla")
data("unemployment_ltla")

uc_children_df <- UC_households_ltla |>
  left_join(children_low_income_ltla) |>
  left_join(unemployment_ltla) |> 
  select(ltla21_code, ltla21_name, year, UC_households_perc, children_low_income_perc, unemployment_perc) |> 
  filter(!is.na(UC_households_perc) & !is.na(children_low_income_perc))

uc_children_16_20_df <- uc_children_df |> 
  filter(year <= 2020)

print(paste0("Number of unique local authorities: ", length(unique(uc_children_16_20_df$ltla21_code))))

head(uc_children_16_20_df, 10)
```
There is data for a total of 340 local authorities covering 2016 to 2022, leading to 2380 observations across England, Wales and Scotland.

### Initial observation
```{r}
summary_statistics <- uc_children_16_20_df |> 
  group_by(year) |> 
  summarise(
    median_children_low_income = median(children_low_income_perc),
    lower_IQR_children_low_income = quantile(children_low_income_perc, 0.25),
    upper_IQR_children_low_income = quantile(children_low_income_perc, 0.75)
  )

print(summary_statistics)
```

There is a noticeable increase in the proportion of children living in relative low income households. 
The first question is: is this rise in child poverty in some parts of the UK concentrated in areas where more people are claiming UC?

### Scatterplot

**Figure 1. Binned scatterplot of the proportion of household claiming UC and proportion of children living in relative low income households across local authorities (2017)**

```{r}
uc_children_2017_df <- uc_children_df |>
  filter(year == 2017)

cor_value <- round(cor(uc_children_2017_df$UC_households_perc, uc_children_2017_df$children_low_income_perc),2)

ggplot(uc_children_2017_df, aes(x = UC_households_perc, y = children_low_income_perc)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue", linewidth = 0.5) +
  labs(x = "Households on UC (%)", 
       y = "Children in Low Income Households (%)") +
  annotate("text", x=7, y=40, label = paste("Correlation:", cor_value), size = 4, colour = "#636363") +
  theme_minimal()+
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 12)
  )
```

Figure 1 is a binned scatterplot of the proportion of households claiming Universal Credit (x-axis) and the proportion of children living in low income families in 2017. It suggests a strong positive relationship across local authorities. Of course, this correlation could be driven by underlying characteristics of the local authorities. (Reeves & Loopstra, 2021: 9)

### Fixed-Effects Model - Continuous Independent Variable (// Reeves & Loopstra, 2021)

A first baseline fixed effects (time and space) model allows to explore whether this relationship remains even after controlling for time-invariant local authority characteristics and time trends. In a second model, we add a time-varying local authority level control: unemployment rate. 

*Note: Standard errors are clustered for repeated observations within local authorities.*

```{r}
uc_children_FE_baseline <- 
  feols(data = uc_children_16_20_df, 
        children_low_income_perc ~ UC_households_perc | ltla21_code + year,
        cluster = ~ltla21_code)

summary(uc_children_FE_baseline)

# Add unemployment control
uc_children_FE_control <- 
  feols(data = uc_children_16_20_df, 
        children_low_income_perc ~ UC_households_perc + unemployment_perc | ltla21_code + year,
        cluster = ~ltla21_code)

summary(uc_children_FE_control)
```

The results of this first model show that for every 1 percentage point increase in households receiving UC, there is a 0.315 percentage point increase in the proportion of children living in low income families. This relationship is statistically significant at all conventional thresholds. This estimate is not strongly affected, neither in its scale nor in its statistical significance when controlling for unemployment rates. This finding reinforces the claim that the parallel trends assumption more plausible is this situation.

#### Adding an interaction term with duration of rollout

I add a variable for the number of months UC has been rolled out in the local authority. 
This is calculating for each year by doing: 01-01-year - month of the UC rollout.

```{r}
data("uc_first_active_month")

# Join the first active date with main dataset & calculate number of months UC 
# has been active by the end of each year
uc_children_16_20_interaction_df <- uc_children_16_20_df |> 
  left_join(uc_first_active_month, by = "ltla21_code") |> 
  mutate(
    months_active = ifelse(
      year >= year(first_active_date),
      ((year - year(first_active_date)) * 12) + (12 - month(first_active_date) + 1),
      0),
    years_active = ifelse(
      year >= year(first_active_date),
      year - year(first_active_date),
      0)
  ) |> 
  select(-first_active_date)
```

Now let's add the interaction term in the fixed-effects model to see if the effect of the UC rollout is moderated by the number of months Universal Credit has been active in the local authority. Standard errors are still clustered by local authority.

```{r}
uc_children_FE_interaction <- 
  feols(data = uc_children_16_20_interaction_df, 
        children_low_income_perc ~ UC_households_perc * years_active + unemployment_perc | ltla21_code + year,
        cluster = ~ltla21_code)

summary(uc_children_FE_interaction)
```

We can visualise the marginal effects of the interaction (change in the increase in the proportion of children in low income families for every 1ppt increase in the proportion of households on UC), using the interaction term:

```{r}
# Extract coefficients
beta_1 <- coef(uc_children_FE_interaction)["UC_households_perc"]
beta_3 <- coef(uc_children_FE_interaction)["UC_households_perc:years_active"]

# For confidence interval:
# Compute variance of the marginal effect of "proportion of households on UC"
# Marginal effect: b1 + b3.Years
# var(aX + bY) = a^2.var(X) + b^2.var(Y) + 2ab.cov(X.Y)
# In our case: var(β1 + β3.Years) = var(β1) + Years^2.var(β3) + 2.Years.cov(β1.β3)

# Extract variances
# Get the variance-covariance matrix of the model's estimates
vcov_matrix <- vcov(uc_children_FE_interaction)

var_beta_1 <- vcov_matrix["UC_households_perc", "UC_households_perc"]
var_beta_3 <- vcov_matrix["UC_households_perc:years_active", "UC_households_perc:years_active"]
cov_beta_1_beta_3 <- vcov_matrix["UC_households_perc", "UC_households_perc:years_active"]

# Build the dataset (95% confidence interval)
interaction_visualisation <-
  tibble(
    Year = 0:5,
    Marginal_Effect = beta_1 + (beta_3 * Year),
    Variance = var_beta_1 + Year^2 * var_beta_3 + 2 * Year * cov_beta_1_beta_3,
    CI_lower = Marginal_Effect - 1.96 * sqrt(Variance),
    CI_upper = Marginal_Effect + 1.96 * sqrt(Variance)
  )

# Visualise
ggplot(interaction_visualisation, aes(x = Year, y = Marginal_Effect)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.1) +
  theme_minimal() + # This sets a minimal theme for the plot
  labs(
    x = "Number of years UC has been active in the local authority",
    y = "Change in proportion of children in poverty for every year's increase in UC"
  ) +
  annotate("text", x = 1, y = 0.48, label = paste("Interaction estimate:", round(beta_3, 3)), size = 5, colour = "#636363") +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 13)
  )

```

### Fixed-Effects Model - Binary Independent Variable (UC Full Service Rollout) (// Hardie, 2023)

In the next model, the Universal Credit rollout is studied as a binary treatment across all local authorities. This additional analysis is critical to assess a causality link of the rollout of the policy on the proportion of children living in low income families because it allows to run placebo tests, which will be detailed later. 

```{r}
data("uc_rollout_ltla_15mo")
data("uc_rollout_ltla_12mo")
data("uc_rollout_ltla_6mo")

uc_rollout_ltla_15mo_joined <- uc_rollout_ltla_15mo |> 
  left_join(children_low_income_ltla)

uc_rollout_ltla_12mo_joined <- uc_rollout_ltla_12mo |> 
  left_join(children_low_income_ltla)

uc_rollout_ltla_6mo_joined <- uc_rollout_ltla_6mo |> 
  left_join(children_low_income_ltla)
```

```{r}
uc_children_binary_15mo_fem <- 
  feols(data = uc_rollout_ltla_15mo_joined, 
        children_low_income_perc ~ uc_rolled_out | ltla21_code + year,
        cluster = ~ltla21_code)

summary(uc_children_binary_15mo_fem)

uc_children_binary_12mo_fem <- 
  feols(data = uc_rollout_ltla_12mo_joined, 
        children_low_income_perc ~ uc_rolled_out | ltla21_code + year,
        cluster = ~ltla21_code)

summary(uc_children_binary_12mo_fem)

uc_children_binary_6mo_fem <- 
  feols(data = uc_rollout_ltla_6mo_joined, 
        children_low_income_perc ~ uc_rolled_out | ltla21_code + year,
        cluster = ~ltla21_code)

summary(uc_children_binary_6mo_fem)

```

None of the models displays a statistically significant effect of the rollout of Universal Credit on the proportion of children in low income families within local authorities. The implications of this finding are difficult to flesh out because of the potential measurement issues emerging from the difference in measurement frequency between the independent variable and outcome. 

