---
title: "The Effects of Welfare Reform on Child Poverty in the UK"
author: "Matteo Larrode"
date: "`r Sys.Date()`"
output: md_document
---
# The Effects of Welfare Reform on Child Poverty in the UK

In this dissertation, I conduct a quantitative analysis of the effect of welfare policy changes in the UK on the proportion of children living in relative low-income families. 

This Markdown includes the data analysis component of this dissertation. It will first include some exploratory data analysis, and then will be used to run the other components of the quantitative analysis of the impact of UK benefit policies on the number of children living in relative low income families. Note that the collection and cleaning of the data that will be used here is done in scripts in the `data-raw` folder of this repository, and the datasets are available in the `data` folder. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r package-loading, include=FALSE}
library(tidyverse)

# pkgload::load_all(".")
```

## Data

### Outcome: Children Living in Relative Low Income Families

Data is downloaded from [Stat-Xplore](https://stat-xplore.dwp.gov.uk/webapi/jsf/login.xhtml), the Department of Work and Pensions (DWP) data distribution platform.

- Low income is a family whose equivalised income is below 60 per cent of median household incomes. For Absolute low income involves, the median of the 2010/11 year is used. For Relative low income, the comparison is made to the median of the current year. This dissertation focuses on Relative low income because it better adjusts for economic changes. Relative income is Before Housing Costs (BHC) and includes contributions from earning, state support and pensions. Equivalisation adjusts incomes for household size and composition.

- A family must have claimed Child Benefit and at least one other household benefit (Universal Credit, tax credits or Housing Benefit) at any point in the year to be classed as low income in these statistics

- Children are defined as dependent individuals aged under 16; or aged 16 to 19 in full-time non-advanced education.

Visit the dedicated [DWP page](https://www.gov.uk/government/publications/children-in-low-income-families-local-area-statistics-background-information-and-methodology/background-information-and-methodology-children-in-low-income-families-local-area-statistics) for further information. 

#### Relative measure

To get the **proportion** of children in low income families, I used [yearly population estimates](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/estimatesofthepopulationforenglandandwales) provided by the Office for National Statistics (for England and Wales), and [time series data](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/population/population-estimates/mid-year-population-estimates/population-estimates-time-series-data) from the National Records of Scotland. Because not all individuals aged 16 to 19 years old are counted as children in the CiRLIF measure, population counts are summed for individuals aged 0-15 years old, and used to divide the absolute measure of CiRLIF provided by the DWP.

**Geography: Local authorities (called ltla for Lower Tier Local Authorities)**

- England: English local authority districts (309)
- Wales: Unitary authorities (22)
- Scotland: Scottish council areas (32)
- Northern Ireland: District council areas (11)

### Main Independent Variable: Households on Universal Credit

Data for the number of households who have a calculated entitlement for Universal Credit is downloaded from [Stat-Xplore](https://stat-xplore.dwp.gov.uk/webapi/jsf/login.xhtml), the Department of Work and Pensions (DWP) data distribution platform.

Data is also available regarding the family type, number and age of children, and work status of the household claiming Universal Credit. For further information, visit the dedicated [DWP page](https://www.gov.uk/government/publications/universal-credit-statistics-background-information-and-methodology/universal-credit-statistics-background-information-and-methodology).

#### Relative measure

To calculate the proportion of households on Universal Credit for each local authority, numbers of households on UC were divided by historical annual estimates of the number of households published by the Office for National Statistics: [2004-2019](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/families/adhocs/13586estimatednumberofhouseholdsbyselectedhouseholdtypeslocalauthoritiesinenglandandwalescountiesandregionsofenglandscottishcouncilareasandgreatbritainconstituentcountries2004to2019), [2020](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/families/adhocs/14432estimatednumberofhouseholdsingreatbritaininnuts1nuts3englishandwelshlocalauthoritiesandscottishcouncilareas2020), [2021](https://www.ons.gov.uk/datasets/TS041/editions/2021/versions/3).

**Geography: Local authorities (called ltla for Lower Tier Local Authorities)**

- England: English local authority districts (309)
- Wales: Unitary authorities (22)
- Scotland: Scottish council areas (32)
- *[Data for Northern Ireland](https://www.communities-ni.gov.uk/articles/universal-credit-statistics) is made available by the Northern Ireland Department for Communities, but not yet collected in this package*

##  I) Relationship Between Households on UC and Children in Low Income Families

I start with an examination of the relationship between the rollout of Universal Credit, more precisely the proportion of households on Universal Credit, and the proportion of children living in low income household. 

I join the datasets of households on UC and children living in low income, on local authority and year.

```{r message=FALSE}
data("children_low_income_ltla")
data("UC_households_ltla")

UC_Children_df <- UC_households_ltla |>
  left_join(children_low_income_ltla) |>
  select(ltla21_code, ltla21_name, year, UC_households_perc, children_low_income_perc) |> 
  filter(!is.na(UC_households_perc) & !is.na(children_low_income_perc))

print(paste0("Number of unique local authorities: ", length(unique(UC_Children_df$ltla21_code))))

head(UC_Children_df, 10)
```
There is data for a total of 340 local authorities covering 2016 to 2022, leading to 2380 observations across England, Wales and Scotland.

### Initial observation
```{r}
summary_statistics <- UC_Children_df |> 
  group_by(year) |> 
  summarise(
    median_children_low_income = median(children_low_income_perc),
    lower_IQR_children_low_income = quantile(children_low_income_perc, 0.25),
    upper_IQR_children_low_income = quantile(children_low_income_perc, 0.75)
  )

print(summary_statistics)
```

There is a noticeable increase in the proportion of children living in relative low income households. 
The first question is: is this rise in child poverty in some parts of the UK concentrated in areas where more people are claiming UC?

### Scatterplot

**Figure 1. Binned scatterplot of the proportion of household claiming UC and proportion of children living in relative low income households across local authorities (2017)**

```{r}
UC_Children_2017_df <- UC_Children_df |>
  filter(year == 2017)

ggplot(UC_Children_2017_df, aes(x = UC_households_perc, y = children_low_income_perc)) +
  geom_point()
```

Figure 1 is a binned scatterplot of the proportion of households claiming Universal Credit (x-axis) and the proportion children living in low income households in 2017. It suggests a strong positive relationship across local authorities. Of course, this correlation could be driven by underlying characteristics of the local authorities. (Reeves & Loopstra, 2021: 9)

### Regression Models 

```{r}
UC_Children_model <- lm(data = UC_Children_df, children_low_income_perc ~ UC_households_perc)
summary(UC_Children_model)
```