---
title: "The Effects of Welfare Reform on Child Poverty in the UK"
author: "Matteo Larrode"
date: "`r Sys.Date()`"
output: md_document
---
# The Effects of Welfare Reform on Child Poverty in the UK

In this dissertation, I conduct a quantitative analysis of the effect of welfare policy changes in the UK on the proportion of children living in relative low-income families. 

This Markdown includes the data analysis component of this dissertation. It will first include some exploratory data analysis, and then will be used to run the other components of the quantitative analysis of the impact of UK benefit policies on the number of children living in relative low income families. Note that the collection and cleaning of the data that will be used here is done in scripts in the `data-raw` folder of this repository, and the datasets are available in the `data` folder.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r package-loading, include=FALSE}
library(tidyverse)
library(viridis)
library(fixest)
library(corrplot)

# pkgload::load_all(".")
```

## Data

### Outcome: Children Living in Relative Low Income Families

Data is downloaded from [Stat-Xplore](https://stat-xplore.dwp.gov.uk/webapi/jsf/login.xhtml), the Department of Work and Pensions (DWP) data distribution platform.

- Low income is a family whose equivalised income is below 60 per cent of median household incomes. For Absolute low income involves, the median of the 2010/11 year is used. For Relative low income, the comparison is made to the median of the current year. This dissertation focuses on Relative low income because it better adjusts for economic changes. Relative income is Before Housing Costs (BHC) and includes contributions from earning, state support and pensions. Equivalisation adjusts incomes for household size and composition.

- A family must have claimed Child Benefit and at least one other household benefit (Universal Credit, tax credits or Housing Benefit) at any point in the year to be classed as low income in these statistics

- Children are defined as dependent individuals aged under 16; or aged 16 to 19 in full-time non-advanced education.

Visit the dedicated [DWP page](https://www.gov.uk/government/publications/children-in-low-income-families-local-area-statistics-background-information-and-methodology/background-information-and-methodology-children-in-low-income-families-local-area-statistics) for further information. 

#### Relative measure

To get the **proportion** of children in low income families, I used [yearly population estimates](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/estimatesofthepopulationforenglandandwales) provided by the Office for National Statistics (for England and Wales), and [time series data](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/population/population-estimates/mid-year-population-estimates/population-estimates-time-series-data) from the National Records of Scotland. Because not all individuals aged 16 to 19 years old are counted as children in the CiRLIF measure, population counts are summed for individuals aged 0-15 years old, and used to divide the absolute measure of CiRLIF provided by the DWP.

**Geography: Local authorities (called ltla for Lower Tier Local Authorities)**

- England: English local authority districts (309)
- Wales: Unitary authorities (22)
- Scotland: Scottish council areas (32)
- Northern Ireland: District council areas (11)

### Main Independent Variable: Households on Universal Credit

Data for the number of households who have a calculated entitlement for Universal Credit is downloaded from [Stat-Xplore](https://stat-xplore.dwp.gov.uk/webapi/jsf/login.xhtml), the Department of Work and Pensions (DWP) data distribution platform.

Data is also available regarding the family type, number and age of children, and work status of the household claiming Universal Credit. For further information, visit the dedicated [DWP page](https://www.gov.uk/government/publications/universal-credit-statistics-background-information-and-methodology/universal-credit-statistics-background-information-and-methodology).

#### Relative measure

To calculate the proportion of households on Universal Credit for each local authority, numbers of households on UC were divided by historical annual estimates of the number of households published by the Office for National Statistics: [2004-2019](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/families/adhocs/13586estimatednumberofhouseholdsbyselectedhouseholdtypeslocalauthoritiesinenglandandwalescountiesandregionsofenglandscottishcouncilareasandgreatbritainconstituentcountries2004to2019), [2020](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/families/adhocs/14432estimatednumberofhouseholdsingreatbritaininnuts1nuts3englishandwelshlocalauthoritiesandscottishcouncilareas2020), [2021](https://www.ons.gov.uk/datasets/TS041/editions/2021/versions/3).

**Geography: Local authorities (called ltla for Lower Tier Local Authorities)**

- England: English local authority districts (309)
- Wales: Unitary authorities (22)
- Scotland: Scottish council areas (32)
- *[Data for Northern Ireland](https://www.communities-ni.gov.uk/articles/universal-credit-statistics) is made available by the Northern Ireland Department for Communities, but not yet collected in this package*

#### Universal Credit Rollout: Binary variable (removed)

The official [UC 'Full Service' transition rollout schedule](https://assets.publishing.service.gov.uk/media/5ab507c8e5274a1aa2d414d1/universal-credit-transition-rollout-schedule.pdf) published by the DWP was used to create explanatory variables measuring cross-local authority variation in the timing of the UC rollout (Hardie, 2023: 1035). The first is a binary variable indicating whether UC ‘Full Service’ had rolled out yet in each year in each local authority in the sample (coded: 0 = ‘no', 1 = ‘yes’). To mitigate bias and ensure robustness in the conversion of the monthly Universal Credit rollout data to annual data for the analysis, I coded multiple variables with different thresholds — December of the previous year, April, and October — to determine the minimum duration of Universal Credit presence within a financial year required for a local authority to be classified as having implemented UC. This first binary variable allows to study the staggered rollout of UC as a natural experiment, with a delayed treatment across local authorities.

The 'Full Service' iteration of Universal Credit was chosen for analysis because it represents the comprehensive form of UC, making it accessible to all types of claimants, including those with housing expenses. The initial 'Live Service' variant was not considered in this study due to its limited scope, primarily involving claims from single, unemployed individuals without housing cost claims, thereby having minimal impact on children from low-income families. (Hardie, 2023: 1035). 

#### Universal Credit Rollout Duration: Categorical variable (needs updating)

If a significant effect of the rollout on children living in low income families in indeed observed, this effect could be expected to vary depending on how long UC has been rollout out in the local authority. One possible scenario is that the effect fades over time because the administrative system adapts. Conversely, as time goes on, more people are reached by Universal Credit, and the effect on children in low income families could become larger. For the purpose of studying the long-term effect of the rollout, we build a secondary independent variable ‘UC Full Service Rollout by Rollout Length’; a categorical variable indicating whether UC ‘Full Service’ had rolled out yet in each month in each local authority , and if so for how long (coded: 0 = ‘No (pre-rollout)’ 1 = ‘Yes (1-6 months post)’, 2 = ‘Yes (7-12 months post)’, 3 = ‘Yes (13-24 months post)’, and 4 = 'Yes(25+ months post)') (Hardie, 2023: 1035). The proportion of children in low income families is measured for a financial year, so the number of months is measured from the month of the rollout to the middle of the financial year (October). 

## Quasi-Randomness of the UC Rollout

The UC ‘Full Service’ rollout was staggered, being introduced in different local authorities at different times between 2016 and 2018 (DWP, 2018). This section examines, both using information released by the DWP, and descriptive analysis, the extent to which the UC rollout can be considered as a form of natural experiment.

Let's start with mapping our pre-treatment variables, and compare them with the years of the rollout of Universal Credit. A function was created in this package to map specified variables. 

```{r message=FALSE}
# Dataset of pre-treatment variables
pkgload::load_all(".")
data(pre_treatment_df)

# Map of quarter of UC rollout
map_variable_ltla(pre_treatment_df, "first_active_quarter", factor = TRUE)

# Map of pre-treatment proportion in children in low income families
map_variable_ltla(pre_treatment_df, "children_low_income_perc")

# Map of pre-treatment unemployment
map_variable_ltla(pre_treatment_df, "unemployment_perc")

# Map of pre-treatment proportion of lone parent households
map_variable_ltla(pre_treatment_df, "lone_parent_households_perc")
```

Now let's examine the pre-treatment socio-demographic characteristics of local authorities, grouped by the quarter during which UC was introduced.

```{r}
half_year_averages <- pre_treatment_df |> 
   mutate(
    year = substr(first_active_quarter, 1, 4), # Extract year part from the quarter info
    quarter = substr(first_active_quarter, 6, 7), # Extract quarter part
    half_year_group = paste(year, ifelse(quarter %in% c("Q1", "Q2"), "H1", "H2"))
  ) |> 
  group_by(half_year_group) |> 
  summarise(
    Number_LAs = n(),
    Avg_Children_Low_Income = mean(children_low_income_perc, na.rm = TRUE),
    Avg_Unemployment = mean(unemployment_perc, na.rm = TRUE),
    Avg_Lone_Parents = mean(lone_parent_households_perc, na.rm = TRUE),
    Avg_Households = mean(households_number, na.rm = TRUE),
    .groups = 'drop'
  )

# Calculate overall averages to add as a total row
overall_averages <- pre_treatment_df |> 
  summarise(
    half_year_group = "Total",
    Number_LAs = n(),
    Avg_Children_Low_Income = mean(children_low_income_perc, na.rm = TRUE),
    Avg_Unemployment = mean(unemployment_perc, na.rm = TRUE),
    Avg_Lone_Parents = mean(lone_parent_households_perc, na.rm = TRUE),
    Avg_Households = mean(households_number, na.rm = TRUE)
  )

balance_table <- rbind(half_year_averages, overall_averages)
balance_table
```

This balance table gives strong support in favour of the consideration of the UC rollout as quasi-random. 

##  I) Relationship Between Households on UC and Children in Low Income Families

I start with an examination of the relationship between the rollout of Universal Credit, more precisely the proportion of households on Universal Credit, and the proportion of children living in low income household. 

I join the datasets of households on UC and children living in low income, on local authority and year.

```{r message=FALSE}
data("dataset_part1")

print(paste0("Number of unique local authorities: ", 
             length(unique(dataset_part1$ltla21_code))))

summary(dataset_part1)
sd(dataset_part1$UC_households_perc)
sd(dataset_part1$children_low_income_perc)
sd(dataset_part1$unemployment_perc)
```
There is data for a total of 339 local authorities covering 2016 to 2020, leading to 1691 observations across England, Wales and Scotland.

### Initial observation
```{r}
summary_statistics <- dataset_part1 |> 
  group_by(year) |> 
  summarise(
    median_children_low_income = median(children_low_income_perc),
    lower_IQR_children_low_income = quantile(children_low_income_perc, 0.25),
    upper_IQR_children_low_income = quantile(children_low_income_perc, 0.75)
  )

print(summary_statistics)
```

There is a noticeable increase in the proportion of children living in relative low income households. 
The first question is: is this rise in child poverty in some parts of the UK concentrated in areas where more people are claiming UC?

### Scatterplot

**Figure 1. Binned scatterplot of the proportion of household claiming UC and proportion of children living in relative low income households across local authorities (2017)**

```{r}
uc_children_2017_df <- dataset_part1 |>
  filter(year == 2017)

cor_value <- round(cor(uc_children_2017_df$UC_households_perc, uc_children_2017_df$children_low_income_perc),2)

ggplot(uc_children_2017_df, aes(x = UC_households_perc, y = children_low_income_perc)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue", linewidth = 0.5) +
  labs(x = "Households on UC (%)", 
       y = "Children in Low Income Households (%)") +
  annotate("text", x=7, y=40, label = paste("Correlation:", cor_value), size = 4, colour = "#636363") +
  theme_minimal()+
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 12)
  )
```

Figure 1 is a binned scatterplot of the proportion of households claiming Universal Credit (x-axis) and the proportion of children living in low income families in 2017. It suggests a strong positive relationship across local authorities. Of course, this correlation could be driven by underlying characteristics of the local authorities. (Reeves & Loopstra, 2021: 9)

### Fixed-Effects Model - Continuous Independent Variable

A first baseline fixed effects (time and space) model allows to explore whether this relationship remains even after controlling for time-invariant local authority characteristics and time trends. In a second model, we add a time-varying local authority level control: unemployment rate. 

*Note: Standard errors are clustered for repeated observations within local authorities.*

```{r}
fe_model_baseline <- 
  feols(data = dataset_part1, 
        children_low_income_perc ~ UC_households_perc | ltla21_code + year,
        cluster = ~ltla21_code)

summary(fe_model_baseline)

# Add unemployment control
fe_model1 <- 
  feols(data = dataset_part1, 
        children_low_income_perc ~ UC_households_perc + unemployment_perc | ltla21_code + year,
        cluster = ~ltla21_code)

summary(fe_model1)

# Weird negative coefficient for unemployment, could there be an interaction?
fe_model_extra <- 
  feols(data = dataset_part1, 
        children_low_income_perc ~ 
          unemployment_perc * UC_households_perc | ltla21_code + year,
        cluster = ~ltla21_code)

summary(fe_model_extra)
```

The results of this first model show that for every 1 percentage point increase in households receiving UC, there is a 0.315 percentage point increase in the proportion of children living in low income families. This relationship is statistically significant at all conventional thresholds. This estimate is not strongly affected, neither in its scale nor in its statistical significance when controlling for unemployment rates. This finding reinforces the claim that the parallel trends assumption holds is this situation.

For context, we can estimate the increase in the absolute number of children living in low income households, in a local authority with the average number of children. For that, we need: 

- to calculate the average percentage point increase of households on UC from the previous year,
- to multiply that by 0.29 to get the estimated percentage point increase in the proportion of children in low income families,
- to get the average number of children each year. 

```{r}
dataset_part1_cont <- UC_households_ltla |>
  left_join(children_low_income_ltla) |>
  left_join(unemployment_ltla) |>
  filter(year <= 2020) |>
  filter(!is.na(UC_households_perc) & !is.na(children_low_income_perc) & !is.na(unemployment_perc)) |>
  select(
    ltla21_code, ltla21_name,
    year, UC_households_perc,
    population_0_16, children_low_income_perc
  )

# Calculate the year-to-year difference in UC_households_perc for each LA
dataset_part1_cont <- dataset_part1_cont |> 
  arrange(ltla21_code, year) |> 
  group_by(ltla21_code) |> 
  mutate(UC_households_perc_diff = UC_households_perc - lag(UC_households_perc),
         children_low_income_perc_diff = children_low_income_perc - lag(children_low_income_perc)) |> 
  ungroup()

yearly_estimates_df <- dataset_part1_cont |> 
  group_by(year) |> 
  summarise(avg_UC_households_perc_diff = mean(UC_households_perc_diff, na.rm = TRUE),
            avg_children_low_income_perc_diff = mean(children_low_income_perc_diff, na.rm = TRUE),
            avg_0_16_pop = mean(population_0_16, na.rm = TRUE)) |> 
  mutate(est_increase_children_perc = avg_UC_households_perc_diff * 0.29,
         est_increase_children_abs = est_increase_children_perc * avg_0_16_pop / 100)
```


## II) Adding an interaction term with duration of rollout

I add a variable for the number of years UC has been rolled out in the local authority. 
This is calculating for each year by doing: 01-01-year - month of the UC rollout.

```{r}
data("dataset_part2")

summary(dataset_part2)
sd(dataset_part2$years_active, na.rm = TRUE)
```

Now let's add the interaction term in the fixed-effects model to see if the effect of the UC rollout is moderated by the number of years Universal Credit has been active in the local authority. Standard errors are still clustered by local authority.

```{r}
fe_model2 <- 
  feols(data = dataset_part2, 
        children_low_income_perc ~ 
          UC_households_perc + 
          years_active +
          UC_households_perc * years_active +
          unemployment_perc | ltla21_code + year,
        cluster = ~ltla21_code)

summary(fe_model2)
```

We can visualise the marginal effects of the interaction (change in the increase in the proportion of children in low income families for every 1ppt increase in the proportion of households on UC), using the interaction term:

```{r}
# Extract coefficients (fe_model2 is the fixed effects model including interaction term with years of UC active)
beta_1 <- coef(fe_model2)["UC_households_perc"]
beta_3 <- coef(fe_model2)["UC_households_perc:years_active"]

# For confidence interval:
# Compute variance of the marginal effect of "proportion of households on UC"
# Marginal effect: b1 + b3.Years
# var(aX + bY) = a^2.var(X) + b^2.var(Y) + 2ab.cov(X.Y)
# In our case: var(β1 + β3.Years) = var(β1) + Years^2.var(β3) + 2.Years.cov(β1.β3)

# Extract variances
# Get the variance-covariance matrix of the model's estimates
vcov_matrix <- vcov(fe_model2)

var_beta_1 <- vcov_matrix["UC_households_perc", "UC_households_perc"]
var_beta_3 <- vcov_matrix["UC_households_perc:years_active", "UC_households_perc:years_active"]
cov_beta_1_beta_3 <- vcov_matrix["UC_households_perc", "UC_households_perc:years_active"]

# Build the dataset (95% confidence interval)
interaction_visualisation <-
  tibble(
    Year = 0:5,
    Marginal_Effect = beta_1 + (beta_3 * Year),
    Variance = var_beta_1 + Year^2 * var_beta_3 + 2 * Year * cov_beta_1_beta_3,
    CI_lower = Marginal_Effect - 1.96 * sqrt(Variance),
    CI_upper = Marginal_Effect + 1.96 * sqrt(Variance)
  )
```

Now on with the visualisation:

```{r}
# Visualise
ggplot(interaction_visualisation, aes(x = Year, y = Marginal_Effect)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.1) +
  scale_y_continuous(limits = c(0, 0.6)) +
  theme_minimal() + # This sets a minimal theme for the plot
  labs(
    x = "Number of years UC has been active in the local authority",
    y = "Change in proportion of children in poverty\n for every 1ppt increase in the proportion of\n households on UC"
  ) +
  annotate("text", x = 1.2, y = 0.55, label = paste("Interaction estimate:", round(beta_3, 3)), size = 6, colour = "#636363") +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 13),
    axis.text = element_text(size =11)
  )

```

## III) Impact of family type on the effect of UC rollout on child poverty  

The literature suggests specific design features of Universal Credit may have affected lone parents disproportionately. 

A first way to examine this possibility is to take into account local authority level (time-varying) characteristics. After having compiled data on the percentage of lone parents households in local authorities (2016-2019), I add an interaction term to the model.

```{r}
data("dataset_part3")

dataset_part3 <- dataset_part3 |> 
  filter(!is.na(lone_parent_households_perc))

summary(dataset_part3$lone_parent_households_perc)
sd(dataset_part3$lone_parent_households_perc, na.rm = TRUE)
```



```{r}
# Baseline with lone parents
fe_model_baseline2 <- 
  feols(data = dataset_part3, 
        children_low_income_perc ~ 
          UC_households_perc + 
          lone_parent_households_perc +
          unemployment_perc | ltla21_code + year,
        cluster = ~ltla21_code)

summary(fe_model_baseline2)


fe_model3 <- 
  feols(data = dataset_part3, 
        children_low_income_perc ~ 
          UC_households_perc + 
          lone_parent_households_perc +
          UC_households_perc * lone_parent_households_perc +
          unemployment_perc | ltla21_code + year,
        cluster = ~ltla21_code)

summary(fe_model3)
```

We note that, when adding the interaction term, the coefficient for `UC_households_perc` becomes negative. This is because of a change in its interpretation.  In the presence of interaction terms, the main effect (coefficient of `UC_households_perc` when the interaction term is included) represents the effect of the variable at the reference level (typically zero) of the interacting variable (`lone_parent_households_perc`). Without the interaction term, the coefficient of `UC_households_perc` represents the overall, or average, effect across all levels of `lone_parent_households_perc`.

Let's visualise this interaction term.

```{r}
# Summary of the values for the proportions of lone parent households
summary(dataset_part3$lone_parent_households_perc)

# Extract coefficients
beta_1_part3 <- coef(fe_model3)["UC_households_perc"]
beta_3_part3 <- coef(fe_model3)["UC_households_perc:lone_parent_households_perc"]

# Extract variances
# Get the variance-covariance matrix of the model's estimates
vcov_matrix_part3 <- vcov(fe_model3)

var_beta_1_part3 <- vcov_matrix_part3["UC_households_perc", "UC_households_perc"]
var_beta_3_part3 <- vcov_matrix_part3["UC_households_perc:lone_parent_households_perc", "UC_households_perc:lone_parent_households_perc"]
cov_beta_1_beta_3_part3 <- vcov_matrix_part3["UC_households_perc", "UC_households_perc:lone_parent_households_perc"]

# Build the dataset (95% confidence interval)
interaction_visualisation_part3 <-
  tibble(
    Lone_Parents = 1:22,
    Marginal_Effect = beta_1_part3 + (beta_3_part3 * Lone_Parents),
    Variance = var_beta_1_part3 + Lone_Parents^2 * var_beta_3_part3 + 2 * Lone_Parents * cov_beta_1_beta_3_part3,
    CI_lower = Marginal_Effect - 1.96 * sqrt(Variance),
    CI_upper = Marginal_Effect + 1.96 * sqrt(Variance)
  )
```

Now on with the visualisation:

```{r}
# Visualise
ggplot(interaction_visualisation_part3, aes(x = Lone_Parents, y = Marginal_Effect)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.1) +
  theme_minimal() + # This sets a minimal theme for the plot
  labs(
    x = "Proportion of lone parent households in the local authority",
    y = "Change in proportion of children in poverty\n for every 1ppt increase in the proportion of\n households on UC"
  ) +
  annotate("text", x = 5, y = 0.5, label = paste("Interaction estimate:", round(beta_3_part3, 3)), size = 7, colour = "#636363") +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 13)
  )

```

Another method to investigate the differentiated impact of UC on lone parent households, rather than to consider family types as a local authority level independent variable, is to consider to model two different dependent variables using data released by the DWP. 

Indeed, the number of children living in low income households is available for different family types. 


## Robustness and sensitivity analysis

### Linear regression analysis assumptions

First, let's check the assumptions underlying regression analysis. 

```{r}
residuals_model3 <- data.frame(residuals(fe_model3)) |> 
  rename(residuals = "residuals.fe_model3.") |> 
  mutate(standardised_residuals = scale(residuals))
```

**Linearity**

```{r}
predicted_values <- predict(fe_model3)

# Plot observed vs. predicted values
plot(dataset_part3$children_low_income_perc, predicted_values, main = "Observed vs. Predicted",
     xlab = "Observed values", ylab = "Predicted values")
abline(0, 1)
```
This is a scatter plot of the observed values against the predicted values from the final model. Ideally, if the model's predictions are perfect, all points would lie on the 45-degree line where the predicted values equal the observed values. In this plot, the points seem to lie around a line that could be close to this ideal. This suggests that there is a linear relationship between the predictors and the response variable, which is a good sign.

However, it's important to note that as the observed values increase, the spread of the predictions also seems to increase slightly. This could be a hint of heteroscedasticity, which is a violation of one of the regression assumptions, and it suggests that the variance of the errors may not be constant across all levels of the predictors.

**Homoscedasticity**

```{r}
# Residuals vs. Fitted values
plot(predicted_values, residuals(fe_model3), main = "Residuals vs. Fitted",
     xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red")
```

This plot shows the residuals (the differences between observed and predicted values) plotted against the fitted (predicted) values. For this plot, a good sign is a random scatter of points, without any discernible pattern, centered around the horizontal line at zero. This would indicate that the residuals have constant variance (homoscedasticity) and are independent of the fitted values.

In this plot once again, while the residuals are centered around the horizontal line at zero (which is good), there seems to be a pattern where the spread of the residuals increases with the fitted values. This is once again an indication of heteroscedasticity. 

Both plots together suggest that while there is a linear relationship between the predictors and the response variable, the assumption of homoscedasticity may not be fully satisfied. It might be helpful to also conduct a formal test for homoscedasticity, such as the Breusch-Pagan test, and consider transforming the response variable or using a different model that accounts for the heteroscedastic nature of the data, such as generalized least squares or robust standard errors.

**Normality of residuals**

```{r}
# QQ plot for residuals
qqnorm(residuals(fe_model3))
qqline(residuals(fe_model3), col = "red")
```

The QQ (Quantile-Quantile) plot compares the quantiles of the observed residuals with the quantiles expected under a normal distribution. The plot displayed a fairly consistent alignment of the data points with the 45-degree reference line in the central section, indicating an approximate adherence to normality within this range. However, the tails of the distribution, particularly the right tail, exhibit a noticeable departure from the reference line (in a "S" shape), suggesting the presence of heavier tails than those typically observed in a normal distribution.

**Independence of residuals**

**Absence of multicollinearity**

Let's start with a correlation matrix. 

```{r}
# Create a correlation matrix of the predictor variables.
cor_matrix <- cor(dataset_part3[, c("UC_households_perc", "lone_parent_households_perc", "unemployment_perc")])

# Plot the correlation matrix
corrplot(cor_matrix, method = "number", type = "upper")
```

- UC_households_perc and lone_parent_households_perc: The correlation coefficient is 0.19, indicating a weak positive relationship. This means that as the percentage of UC households increases, there is a slight tendency for the percentage of lone parent households to increase as well, but the relationship is not strong.

- UC_households_perc and unemployment_perc: Similarly, there is a correlation coefficient of 0.19, which also indicates a weak positive relationship. The percentage of UC households and the unemployment percentage tend to increase together, but again, the relationship is not pronounced.

- lone_parent_households_perc and unemployment_perc: The correlation coefficient here is 0.52, suggesting a moderate positive relationship. As the percentage of lone parent households increases, the unemployment percentage tends to increase more noticeably. This is a more substantial correlation compared to the other pairs but still not so high as to typically indicate a severe risk of multicollinearity on its own.

Based on these correlation coefficients, it does not appear that there is a high risk of multicollinearity between these variables. Multicollinearity is usually of concern when correlation coefficients are above 0.7 or 0.8.

### Outliers

The residuals vs. fitted plot does indicate the presence of potential outliers, as some residuals are quite far from the main cluster centered around the horizontal line at zero. Let us detect them:

```{r}
dataset_part3$residuals <- residuals(fe_model3)
dataset_part3$residuals_standardised <- scale(dataset_part3$residuals)[,1]
```

Unfortunately, it no 'hatvalues' (leverage) method has been implemented for estimations with fixed-effects, making it difficult to assess the influence of these outliers on regression results. 

Another solution to evaluate the extent to which results of the model are sensitive to changes in the sample is to conduct a Monte Carlo simulation. This technique could allow to understand the variability in estimates due to sample variation. First, 100 datasets are created by resampling from the original data. The model is fit to each simulated dataset, and we can observe the distribution of the estimates.

### Monte-carlo simulation of coefficients with different samples

```{r}
# Number of simulations
n_sim <- 100

unique_ltlas <- unique(dataset_part3$ltla21_code)
n_ltlas <- length(unique_ltlas)

coefficients_df <- tibble(
  "UC_households_perc" = numeric(), 
  "lone_parent_households_perc" = numeric(),
  "unemployment_perc" = numeric(), 
  "UC_households_perc_x_lone_parent_households_perc" = numeric())

for (i in 1:n_sim) {
  # Sample local authorities with replacement
  sampled_ltlas <- sample(unique_ltlas, size = 250, replace = TRUE)
  
  # Get rows from the dataset corresponding to the sampled local authorities
  sampled_data <- dataset_part3[dataset_part3$ltla21_code %in% sampled_ltlas, ]
  
  # Fit the model to the sampled data
  sim_model <- feols(data = sampled_data, 
                     children_low_income_perc ~ 
                       UC_households_perc + 
                       lone_parent_households_perc +
                       UC_households_perc * lone_parent_households_perc +
                       unemployment_perc | ltla21_code + year,
                     cluster = ~ltla21_code)
  
  # Store the coefficients
  coefficients_df <- coefficients_df |>
    add_row(
      tibble_row(
        UC_households_perc = coef(sim_model)[1],
        lone_parent_households_perc = coef(sim_model)[2],
        unemployment_perc = coef(sim_model)[3],
        UC_households_perc_x_lone_parent_households_perc = coef(sim_model)[4]
        )
      )
}

# Create a density plot of the simulation results for UC_households_perc_x_lone_parent_households_perc
ggplot(coefficients_df, aes(x = UC_households_perc_x_lone_parent_households_perc)) +
  geom_density(fill = "blue", alpha = 0.5) +
  theme_minimal() +
  labs(title = "Density of Interaction Term from Monte Carlo Simulation",
       x = "UC_households_perc * lone_parent_households_perc",
       y = "Density")
```