---
title: "The Effects of Welfare Reform on Child Poverty in the UK"
author: "Matteo Larrode"
date: "`r Sys.Date()`"
output: md_document
---
# The Effects of Welfare Reform on Child Poverty in the UK

In this dissertation, I conduct a quantitative analysis of the effect of welfare policy changes in the UK on the proportion of children living in relative low-income families. 

This Markdown includes the data analysis component of this dissertation. It will first include some exploratory data analysis, and then will be used to run the other components of the quantitative analysis of the impact of UK benefit policies on the number of children living in relative low income families. Note that the collection and cleaning of the data that will be used here is done in scripts in the `data-raw` folder of this repository, and the datasets are available in the `data` folder. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r package-loading, include=FALSE}
library(tidyverse)
library(fixest)

# pkgload::load_all(".")
```

## Data

### Outcome: Children Living in Relative Low Income Families

Data is downloaded from [Stat-Xplore](https://stat-xplore.dwp.gov.uk/webapi/jsf/login.xhtml), the Department of Work and Pensions (DWP) data distribution platform.

- Low income is a family whose equivalised income is below 60 per cent of median household incomes. For Absolute low income involves, the median of the 2010/11 year is used. For Relative low income, the comparison is made to the median of the current year. This dissertation focuses on Relative low income because it better adjusts for economic changes. Relative income is Before Housing Costs (BHC) and includes contributions from earning, state support and pensions. Equivalisation adjusts incomes for household size and composition.

- A family must have claimed Child Benefit and at least one other household benefit (Universal Credit, tax credits or Housing Benefit) at any point in the year to be classed as low income in these statistics

- Children are defined as dependent individuals aged under 16; or aged 16 to 19 in full-time non-advanced education.

Visit the dedicated [DWP page](https://www.gov.uk/government/publications/children-in-low-income-families-local-area-statistics-background-information-and-methodology/background-information-and-methodology-children-in-low-income-families-local-area-statistics) for further information. 

#### Relative measure

To get the **proportion** of children in low income families, I used [yearly population estimates](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/estimatesofthepopulationforenglandandwales) provided by the Office for National Statistics (for England and Wales), and [time series data](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/population/population-estimates/mid-year-population-estimates/population-estimates-time-series-data) from the National Records of Scotland. Because not all individuals aged 16 to 19 years old are counted as children in the CiRLIF measure, population counts are summed for individuals aged 0-15 years old, and used to divide the absolute measure of CiRLIF provided by the DWP.

**Geography: Local authorities (called ltla for Lower Tier Local Authorities)**

- England: English local authority districts (309)
- Wales: Unitary authorities (22)
- Scotland: Scottish council areas (32)
- Northern Ireland: District council areas (11)

### Main Independent Variable: Households on Universal Credit

Data for the number of households who have a calculated entitlement for Universal Credit is downloaded from [Stat-Xplore](https://stat-xplore.dwp.gov.uk/webapi/jsf/login.xhtml), the Department of Work and Pensions (DWP) data distribution platform.

Data is also available regarding the family type, number and age of children, and work status of the household claiming Universal Credit. For further information, visit the dedicated [DWP page](https://www.gov.uk/government/publications/universal-credit-statistics-background-information-and-methodology/universal-credit-statistics-background-information-and-methodology).

#### Relative measure

To calculate the proportion of households on Universal Credit for each local authority, numbers of households on UC were divided by historical annual estimates of the number of households published by the Office for National Statistics: [2004-2019](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/families/adhocs/13586estimatednumberofhouseholdsbyselectedhouseholdtypeslocalauthoritiesinenglandandwalescountiesandregionsofenglandscottishcouncilareasandgreatbritainconstituentcountries2004to2019), [2020](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/families/adhocs/14432estimatednumberofhouseholdsingreatbritaininnuts1nuts3englishandwelshlocalauthoritiesandscottishcouncilareas2020), [2021](https://www.ons.gov.uk/datasets/TS041/editions/2021/versions/3).

**Geography: Local authorities (called ltla for Lower Tier Local Authorities)**

- England: English local authority districts (309)
- Wales: Unitary authorities (22)
- Scotland: Scottish council areas (32)
- *[Data for Northern Ireland](https://www.communities-ni.gov.uk/articles/universal-credit-statistics) is made available by the Northern Ireland Department for Communities, but not yet collected in this package*

#### Universal Credit Rollout: Binary variable

The official [UC 'Full Service' transition rollout schedule](https://assets.publishing.service.gov.uk/media/5ab507c8e5274a1aa2d414d1/universal-credit-transition-rollout-schedule.pdf) published by the DWP was used to create explanatory variables measuring cross-local authority variation in the timing of the UC rollout (Hardie, 2023: 1035). The first is a binary variable indicating whether UC ‘Full Service’ had rolled out yet in each year in each local authority in the sample (coded: 0 = ‘no', 1 = ‘yes’). To mitigate bias and ensure robustness in the conversion of the monthly Universal Credit rollout data to annual data for the analysis, I coded multiple variables with different thresholds — December of the previous year, April, and October — to determine the minimum duration of Universal Credit presence within a financial year required for a local authority to be classified as having implemented UC. This first binary variable allows to study the staggered rollout of UC as a natural experiment, with a delayed treatment across local authorities.

The 'Full Service' iteration of Universal Credit was chosen for analysis because it represents the comprehensive form of UC, making it accessible to all types of claimants, including those with housing expenses. The initial 'Live Service' variant was not considered in this study due to its limited scope, primarily involving claims from single, unemployed individuals without housing cost claims, thereby having minimal impact on children from low-income families. (Hardie, 2023: 1035). 

#### Universal Credit Rollout Duration: Categorical variable

If a significant effect of the rollout on children living in low income families in indeed observed, this effect could be expected to vary depending on how long UC has been rollout out in the local authority. One possible scenario is that the effect fades over time because the administrative system adapts. Conversely, as time goes on, more people are reached by Universal Credit, and the effect on children in low income families could become larger. For the purpose of studying the long-term effect of the rollout, we build a secondary independent variable ‘UC Full Service Rollout by Rollout Length’; a categorical variable indicating whether UC ‘Full Service’ had rolled out yet in each month in each local authority , and if so for how long (coded: 0 = ‘No (pre-rollout)’ 1 = ‘Yes (1-6 months post)’, 2 = ‘Yes (7-12 months post)’, 3 = ‘Yes (13-24 months post)’, and 4 = 'Yes(25+ months post)') (Hardie, 2023: 1035). The proportion of children in low income families is measured for a financial year, so the number of months is measured from the month of the rollout to the middle of the financial year (October). 

##  I) Relationship Between Households on UC and Children in Low Income Families

I start with an examination of the relationship between the rollout of Universal Credit, more precisely the proportion of households on Universal Credit, and the proportion of children living in low income household. 

I join the datasets of households on UC and children living in low income, on local authority and year.

```{r message=FALSE}
data("children_low_income_ltla")
data("UC_households_ltla")

uc_children_df <- UC_households_ltla |>
  left_join(children_low_income_ltla) |>
  select(ltla21_code, ltla21_name, year, UC_households_perc, children_low_income_perc) |> 
  filter(!is.na(UC_households_perc) & !is.na(children_low_income_perc))

print(paste0("Number of unique local authorities: ", length(unique(uc_children_df$ltla21_code))))

head(uc_children_df, 10)
```
There is data for a total of 340 local authorities covering 2016 to 2022, leading to 2380 observations across England, Wales and Scotland.

### Initial observation
```{r}
summary_statistics <- uc_children_df |> 
  group_by(year) |> 
  summarise(
    median_children_low_income = median(children_low_income_perc),
    lower_IQR_children_low_income = quantile(children_low_income_perc, 0.25),
    upper_IQR_children_low_income = quantile(children_low_income_perc, 0.75)
  )

print(summary_statistics)
```

There is a noticeable increase in the proportion of children living in relative low income households. 
The first question is: is this rise in child poverty in some parts of the UK concentrated in areas where more people are claiming UC?

### Scatterplot

**Figure 1. Binned scatterplot of the proportion of household claiming UC and proportion of children living in relative low income households across local authorities (2017)**

```{r}
uc_children_2017_df <- uc_children_df |>
  filter(year == 2017)

ggplot(uc_children_2017_df, aes(x = UC_households_perc, y = children_low_income_perc)) +
  geom_point()
```

Figure 1 is a binned scatterplot of the proportion of households claiming Universal Credit (x-axis) and the proportion of children living in low income families in 2017. It suggests a strong positive relationship across local authorities. Of course, this correlation could be driven by underlying characteristics of the local authorities. (Reeves & Loopstra, 2021: 9)

### Fixed-Effects Model - Continuous Independent Variable (// Reeves & Loopstra, 2021)

A first baseline fixed effects (time and space) model allows to explore whether this relationship remains even after controlling for time-invariant local authority characteristics and time trends.

*Note: Standard errors are clustered for repeated observations within local authorities.*

```{r}
uc_children_cont_fem <- 
  feols(data = uc_children_df, 
        children_low_income_perc ~ UC_households_perc | ltla21_code + year,
        cluster = ~ltla21_code)

summary(uc_children_cont_fem)
```

The results of this first model show that for every 1 percentage point increase in households receiving UC, there is a 0.138 percentage point increase in the proportion of children living in low income families. This relationship is statistically significant at all conventional thresholds. 

**Important notes**:

1. The key assumption needed for a valid estimation of a causal effect with this fixed-effects model is the parallel trends assumption. That is, local authorities which have **not** displayed an increase in UC caseload would have seen the same change in the proportion of children living in low income families as local authorities which have indeed displayed an increase in proportion of households on UC. This assumption needs to be studied further.
2. It would be benefiticial to control for potentially time varying local authority characteristics such as unemployment. This would make the parallel trends assumption more plausible. More data collection is necessary for that.

### Fixed-Effects Model - Binary Independent Variable (UC Full Service Rollout) (// Hardie, 2023)

In the next model, the Universal Credit rollout is studied as a binary treatment across all local authorities. This additional analysis is critical to assess a causality link of the rollout of the policy on the proportion of children living in low income families because it allows to run placebo tests, which will be detailed later. 

```{r}
data("uc_rollout_ltla_15mo")
data("uc_rollout_ltla_12mo")
data("uc_rollout_ltla_6mo")

uc_rollout_ltla_15mo_joined <- uc_rollout_ltla_15mo |> 
  left_join(children_low_income_ltla)

uc_rollout_ltla_12mo_joined <- uc_rollout_ltla_12mo |> 
  left_join(children_low_income_ltla)

uc_rollout_ltla_6mo_joined <- uc_rollout_ltla_6mo |> 
  left_join(children_low_income_ltla)
```

```{r}
uc_children_binary_15mo_fem <- 
  feols(data = uc_rollout_ltla_15mo_joined, 
        children_low_income_perc ~ uc_rolled_out | ltla21_code + year,
        cluster = ~ltla21_code)

summary(uc_children_binary_15mo_fem)

uc_children_binary_12mo_fem <- 
  feols(data = uc_rollout_ltla_12mo_joined, 
        children_low_income_perc ~ uc_rolled_out | ltla21_code + year,
        cluster = ~ltla21_code)

summary(uc_children_binary_12mo_fem)

uc_children_binary_6mo_fem <- 
  feols(data = uc_rollout_ltla_6mo_joined, 
        children_low_income_perc ~ uc_rolled_out | ltla21_code + year,
        cluster = ~ltla21_code)

summary(uc_children_binary_6mo_fem)

```


### Add time-varying controls to fixed-effects model 

- Research on drivers of child poverty needed
  - unemployment rate
  - nth percentile of wages
  
### Fixed-Effects Model - Categorical Variable: Long-term effects of UC Full Service Rollout

- How to count number of months from rollout to reference year
  - e.g. FYE 2018 (April 2017 - March 2018): if UC has been rolled out in December 2016, how many months are counted for FYE 2018?
  
### Scottish Child Payment 

"On 15 February 2021 we introduced the Scottish Child Payment for low-income families with children under six as part of our work on tackling child poverty. It was initially set at a rate of £10 per week per child with no limit to the number of eligible children.

In March 2022 we published “Best Start, Bright Futures: tackling child poverty delivery plan 2022 to 2026” in which we committed to doubling the value of the Scottish Child Payment from April 2022, delivering the benefit in full to all eligible children under the age of 16 and further increasing the value to £25 per child, per week, by the end of 2022. These commitments were delivered on 14 November 2022." (https://www.gov.scot/policies/social-security/scottish-child-payment/)

- MSOA level children living in low income families for Scotland and England
  - difference in differences for pre-2021 and 2021-2022
  - with matching on UC takeup, unemployment & pre-treatment outcome

### Two-Child Limit in Universal Credit 

- Households with a third or subsequent child born from 6 April 2017 claiming Universal Credit or Child Tax Credit no longer receive additional amounts for these children. The policy only applies to children born from 6 April 2017, so not all families with a third or subsequent child claiming Universal Credit will be affected until the mid-2030s. In April 2023, the two-child limit affected 422,000 (55%) of the 772,000 families with three or more children claiming Universal Credit or Child Tax Credit. 22,000 households had an exception.

- Initial results: Relative poverty among families with three or more children, which has been rising since 2013, has continued to increase since April 2017. The Government points to falling absolute poverty over the period and questions the use of relative poverty measures.

- More research needed (Patrick et al, 2023)

- Possibility of quasi-experimental research? Data on number of children living in low income by age available. 
